---
title: "Analysis of World Development Indicators"
author: "Sissi Li"
date: 2026-02-20
format:
  revealjs:
    theme: moon
    slide-number: true
    transition: fade
    background-transition: zoom
    center: true
    incremental: true
bibliography: references.bib
---

```{r}
df <- read.csv("wdi.csv")
```

## Descriptive Statistics

-   Life expectancy averages around 73 years

-   Unemployment averages approximately 7%

-   GDP per capita is highly right-skewed

-   Large gap between mean and median GDP per capita

```{r}
summary(df$life_expectancy)
summary(df$unemployment_rate)
summary(df$gdp_per_capita)
sd(df$life_expectancy, na.rm = TRUE)
sd(df$unemployment_rate, na.rm = TRUE)
sd(df$gdp_per_capita, na.rm = TRUE)
```

```{r}
cor(df[, c("life_expectancy",
           "unemployment_rate",
           "gdp_per_capita")],
    use = "complete.obs")
```

## GDP per Capita and Life Expectancy

-   Positive association observed

-   Higher-income countries tend to have longer life expectancy

-   Relationship appears nonlinear (diminishing returns at high GDP

```{r}
library(ggplot2)

ggplot(df, aes(x = gdp_per_capita, y = life_expectancy)) +
  geom_point(alpha = 0.6) +
  labs(
    x = "GDP per Capita (USD)",
    y = "Life Expectancy (Years)"
  ) +
  theme_minimal()
```

## Unemployment Patterns by Income Level

-   Countries below mean GDP show greater dispersion

-   Extreme unemployment outliers are more common in lower-income countries

-   Labor market instability appears more pronounced in developing economies

```{r}
#| fig-width: 10
#| fig-height: 6
library(dplyr)

df_clean <- df %>%
  filter(!is.na(unemployment_rate),
         !is.na(gdp_per_capita)) %>%
  mutate(gdp_group = ifelse(gdp_per_capita > mean(gdp_per_capita, na.rm = TRUE),
                            "Above Mean GDP",
                            "Below Mean GDP"))

ggplot(df_clean, aes(x = gdp_group, y = unemployment_rate)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(
    x = "GDP per Capita Group",
    y = "Unemployment Rate (%)"
  ) +
 theme_minimal(base_size = 24)
```

## Summarizing Table

```{r}
library(knitr)

df_clean <- df %>%
  filter(!is.na(life_expectancy),
         !is.na(unemployment_rate),
         !is.na(gdp_per_capita))

summary_table <- data.frame(
  Variable = c("Life Expectancy",
               "Unemployment Rate",
               "GDP per Capita"),
  
  Mean = c(mean(df_clean$life_expectancy),
           mean(df_clean$unemployment_rate),
           mean(df_clean$gdp_per_capita)),
  
  Median = c(median(df_clean$life_expectancy),
             median(df_clean$unemployment_rate),
             median(df_clean$gdp_per_capita)),
  
  `Std. Dev.` = c(sd(df_clean$life_expectancy),
                  sd(df_clean$unemployment_rate),
                  sd(df_clean$gdp_per_capita))
)

kable(summary_table, digits = 2)

```

## Summary

-   Economic development and health outcomes are strongly linked

-   Income distribution across countries is highly unequal

-   Understanding these relationships is essential for policy design

-   Further research could explore causal mechanisms
