---
title: "Analysis of World Development Indicators"
author: "Sissi Li"
date: 2026/2/20
format:
  html: default
  pdf: default
bibliography: references.bib
---

## Q2:Load the Data

```{r}

df <- read.csv("wdi.csv")

```

### Q3: Conduct exploratory data analysis on at least three indicators of your choice. Summarise your findings in markdown sections. Show your code and results.

Indicator 1: life_expectancy
```{r}
summary(df$life_expectancy)
```

Indicator 2: unemployment_rate
```{r}
summary(df$unemployment_rate)
```

Indicator 3: gdp_per_capita
```{r}
summary(df$gdp_per_capita)
```

```{r}
sd(df$life_expectancy, na.rm = TRUE)
sd(df$unemployment_rate, na.rm = TRUE)
sd(df$gdp_per_capita, na.rm = TRUE)
```

```{r}
cor(df[, c("life_expectancy",
           "unemployment_rate",
           "gdp_per_capita")],
    use = "complete.obs")
```

Summary: 
Life expectancy ranges from 18.82 to 85.75 years, with a mean of 73.11 and a median of 74.16, suggesting a fairly symmetric distribution centered in the low-to-mid 70s. The interquartile range (67.79 to 78.53) indicates that most countries cluster within a relatively narrow band. The standard deviation of 7.94 years suggests moderate dispersion across countries. Overall, life expectancy appears relatively stable compared to the other indicators.

Unemployment rates vary substantially, ranging from 0.13% to 36.47%. The mean (7.20%) is higher than the median (5.34%), indicating a right-skewed distribution, likely driven by countries with very high unemployment levels. The standard deviation is 5.85 percentage points, reflecting notable cross-country variation. There are 31 missing observations, which should be considered when interpreting results.

GDP per capita shows extremely wide variation, from $303 to $226,052, with a median of $7,656 and a much larger mean of $21,143. The large gap between the mean and median, combined with a very high standard deviation (31,040), indicates a strong right-skewed distribution, driven by a small number of very wealthy countries. There are 8 missing observations. This degree of dispersion suggests that a log transformation may be useful in further analysis.

### Q4
```{r}
#| fig-cap: "Figure 1: Relationship between GDP per capita and Life Expectancy. Source: World Bank World Development Indicators."
#| label: fig-gdp
#| echo: false

library(ggplot2)

ggplot(df, aes(x = gdp_per_capita, y = life_expectancy)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "GDP per Capita and Life Expectancy",
    x = "GDP per Capita (USD)",
    y = "Life Expectancy (Years)"
  ) +
  theme_minimal()
```

```{r}
#| fig-cap: "Figure 2: Distribution of Unemployment Rates by GDP per Capita Group. Countries are split at the sample mean GDP per capita ($21,143). Source: World Bank World Development Indicators."
#| label: fig-unemp
#| echo: false
#| warning: false
#| message: false

library(dplyr)
library(ggplot2)

# Remove missing values for relevant variables
df_clean <- df %>%
  filter(!is.na(unemployment_rate),
         !is.na(gdp_per_capita))

# Create GDP groups
df_clean <- df_clean %>%
  mutate(gdp_group = ifelse(gdp_per_capita > 21143,
                            "Above Mean GDP",
                            "Below Mean GDP"))

# Boxplot
ggplot(df_clean, aes(x = gdp_group, y = unemployment_rate)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Unemployment Rate by GDP per Capita Group",
    x = "GDP per Capita Group",
    y = "Unemployment Rate (%)"
  ) +
  theme_minimal()

```

### Q5
```{r}
#| tbl-cap: "Summary statistics of key indicators. Source: World Bank World Development Indicators."
#| label: tbl-summary
#| echo: false
      
library(dplyr)
library(knitr)

# Remove rows with NA in any of the three variables
df_clean <- df %>%
  filter(!is.na(life_expectancy),
         !is.na(unemployment_rate),
         !is.na(gdp_per_capita))

# Create summary table
summary_table <- data.frame(
  Variable = c("Life Expectancy",
               "Unemployment Rate",
               "GDP per Capita"),
  
  Mean = c(mean(df_clean$life_expectancy),
           mean(df_clean$unemployment_rate),
           mean(df_clean$gdp_per_capita)),
  
  Median = c(median(df_clean$life_expectancy),
             median(df_clean$unemployment_rate),
             median(df_clean$gdp_per_capita)),
  
  `Std. Dev.` = c(sd(df_clean$life_expectancy),
                  sd(df_clean$unemployment_rate),
                  sd(df_clean$gdp_per_capita)),
  
  Min = c(min(df_clean$life_expectancy),
          min(df_clean$unemployment_rate),
          min(df_clean$gdp_per_capita)),
  
  Max = c(max(df_clean$life_expectancy),
          max(df_clean$unemployment_rate),
          max(df_clean$gdp_per_capita)),
  
  N = c(nrow(df_clean),
        nrow(df_clean),
        nrow(df_clean))
)

kable(summary_table,
      digits = 2,
      caption = "Table 1: Summary Statistics of Key Indicators (NA values removed). Source: World Bank World Development Indicators.")
```

### Q6
As shown in @fig-gdp,, GDP per capita is positively associated with life expectancy.
As shown in @fig-unemp, countries with GDP per capita below the sample mean exhibit greater dispersion in unemployment rates compared to countries above the mean.
Descriptive statistics are presented in @tbl-summary.

### Q7
Mandel and Liebens @mandel2019gdp show that unemployment is related to GDP.
Health also plays a significant role in economic growth [@neofytidou2020health].


